\documentclass[11pt,a4paper]{article}

% ---------- Packages ----------
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{tcolorbox}
\tcbuselibrary{skins,breakable}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{array}
\usepackage{fancyhdr}
\usepackage{titlesec}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,calc}
\usepackage[protrusion=true,expansion=false]{microtype}
\usepackage{siunitx}
\usepackage{caption}

% ---------- Colors ----------
\definecolor{primary}{RGB}{30,58,95}
\definecolor{accent}{RGB}{178,34,52}
\definecolor{softgray}{RGB}{245,245,248}
\definecolor{emerald}{RGB}{27,115,82}
\definecolor{amber}{RGB}{184,115,20}
\definecolor{lightblue}{RGB}{230,240,250}
\definecolor{lightred}{RGB}{253,235,235}
\definecolor{lightgreen}{RGB}{232,245,237}
\definecolor{lightyellow}{RGB}{253,247,230}

% ---------- Hyperref ----------
\hypersetup{
    colorlinks=true,
    linkcolor=primary,
    urlcolor=accent,
    citecolor=primary,
    pdftitle={Lecture 21 -- Performance Metrics},
    pdfauthor={Class notes}
}

% ---------- Titles ----------
\titleformat{\section}
  {\color{primary}\Large\bfseries}{\thesection}{0.6em}{}[{\color{primary}\titlerule[0.8pt]}]
\titleformat{\subsection}
  {\color{accent}\large\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\subsubsection}
  {\color{primary}\normalsize\bfseries}{\thesubsubsection}{0.4em}{}

% ---------- Header/Footer ----------
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small\color{primary}\textbf{Lec 21 --- Performance Metrics}}
\fancyhead[R]{\small\color{primary}Parallel \& Distributed Programming}
\fancyfoot[C]{\small\thepage}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\headrule}{\hbox to\headwidth{\color{primary}\leaders\hrule height \headrulewidth\hfill}}

% ---------- Custom boxes ----------
\newtcolorbox{definitionbox}[1][]{
  enhanced, breakable,
  colback=lightblue, colframe=primary,
  fonttitle=\bfseries\color{white},
  coltitle=white,
  boxrule=0.5pt, arc=3pt,
  left=8pt,right=8pt,top=6pt,bottom=6pt,
  title=#1
}
\newtcolorbox{errorbox}[1][]{
  enhanced, breakable,
  colback=lightred, colframe=accent,
  fonttitle=\bfseries\color{white},
  coltitle=white,
  boxrule=0.5pt, arc=3pt,
  left=8pt,right=8pt,top=6pt,bottom=6pt,
  title=#1
}
\newtcolorbox{insightbox}[1][]{
  enhanced, breakable,
  colback=lightgreen, colframe=emerald,
  fonttitle=\bfseries\color{white},
  coltitle=white,
  boxrule=0.5pt, arc=3pt,
  left=8pt,right=8pt,top=6pt,bottom=6pt,
  title=#1
}
\newtcolorbox{examplebox}[1][]{
  enhanced, breakable,
  colback=lightyellow, colframe=amber,
  fonttitle=\bfseries\color{white},
  coltitle=white,
  boxrule=0.5pt, arc=3pt,
  left=8pt,right=8pt,top=6pt,bottom=6pt,
  title=#1
}
\newtcolorbox{qabox}[1][]{
  enhanced, breakable,
  colback=softgray, colframe=primary,
  fonttitle=\bfseries\color{white},
  coltitle=white,
  boxrule=0.5pt, arc=3pt,
  left=8pt,right=8pt,top=6pt,bottom=6pt,
  title=#1
}

% ---------- Math macros ----------
\newcommand{\Tone}{T_{1}}
\newcommand{\Tp}{T_{p}}
\newcommand{\Sp}{S_{p}}
\newcommand{\Ep}{E_{p}}
\newcommand{\Cp}{C_{p}}
\newcommand{\To}{T_{o}}

% ---------- Title page ----------
\title{
  \vspace{-1.2cm}
  {\color{primary}\Huge\textbf{Performance Metrics}}\\[0.4em]
  {\color{accent}\Large Lecture 21 --- Parallel \& Distributed Programming}\\[0.3em]
  \rule{\linewidth}{0.5pt}\\[-0.2em]
  {\normalsize\itshape Class notes with corrections to slides}
}
\author{\textbf{Lecturer:} Prof. Subodh Sharma \\ \textbf{Institution:} Indian Institute of Technology \\ \textbf{Course:} Introduction to Parallel \& Distributed Programming \\ \textbf{Transcribed by:} Anirudha Saraf, Manvendra Rajpurohit, Gaurav Raju Shende}
\date{}
\linespread{0.95}

\begin{document}
\maketitle
\thispagestyle{empty}

% ---------- Preface ----------
\begin{tcolorbox}[colback=softgray, colframe=primary, boxrule=0.5pt, arc=3pt]
\textbf{About these notes.} These are our notes from Lecture 21 on performance metrics
for parallel programs. We follow the order in which Sir presented the material, fix
a few mistakes in the printed slides (flagged in red boxes), and fill in the
derivations he wrote out on the board but which are not on the deck. The last
section (iso-efficiency and the ``strongly / weakly scalable'' classification)
was given verbally in the closing minutes of class and does not appear on the
slides at all, so we have written it up from notes.
\end{tcolorbox}

\tableofcontents
\newpage

% ==========================================================
\section{Why we care about performance metrics}
% ==========================================================

Correctness is one half of the job. The whole point of writing a parallel program
in the first place is to finish a workload faster, or to handle a bigger workload
than a single machine can, or to use the hardware more economically. None of those
claims mean anything unless we have a way to measure them.

Sir made the point at the start of class that this material could really have been
covered on day one --- the moment you write your first OpenMP, MPI or CUDA program,
you start wanting to know whether the work you put in actually paid off. The answer
has several parts: how does the program behave as you give it more cores, how does
it behave as the problem itself grows, and how much of the runtime is being eaten
by overheads (communication, cache coherence, synchronisation, ordering fences,
and so on).

This lecture pins down the vocabulary --- speedup, cost, overhead, efficiency, and
scalability --- and then presents the two laws (Amdahl and Gustafson) that say
something about how far parallelism can take you. The closing topic, iso-efficiency,
ties everything together by asking how fast the problem must grow if we want to
keep the processors usefully busy.

\begin{insightbox}[The question we are trying to answer]
Given a sequential program and $p$ processors: how much faster does it run, how
well are we using those $p$ processors, and how big does the problem need to be
for that quality to hold up?
\end{insightbox}

% ==========================================================
\section{The RAM model}
% ==========================================================

Before defining anything, Sir fixed a model of computation. Throughout this course
we use the standard \textbf{RAM (Random Access Machine)} model, with these
assumptions:

\begin{itemize}[leftmargin=*]
  \item Every simple arithmetic or logical operation takes one unit of time.
  \item Every local memory access (by a thread or process) takes one unit of time.
  \item Each thread or process is an independent unit of execution with its own local view of memory.
  \item If process $i$ sends a constant-size message to process $j$, the transmission also takes one unit of time.
\end{itemize}

The model is an idealisation --- real hardware has caches, NUMA effects, network
contention, and so on --- but it is the baseline. The communication-aware extensions
later in the lecture refine it without throwing it away.

% ==========================================================
\section{The basic definitions}
% ==========================================================

\subsection{Sequential and parallel time}

Throughout the lecture we write
\begin{itemize}[leftmargin=*]
  \item $\Tone$ for the wall-clock time of the best sequential algorithm for the problem on a single processor, and
  \item $\Tp$ for the wall-clock time of the parallel algorithm on $p$ processors.
\end{itemize}
A small but important point Sir emphasised: $\Tone$ is the time of the
\emph{fastest known serial algorithm}, not the parallel algorithm running on a
single processor. Otherwise we could make a slow parallel algorithm look good
just by comparing it to itself.

\subsection{Speedup}

\begin{definitionbox}[Speedup]
\[
\Sp \;:=\; \frac{\Tone}{\Tp}
\]
$\Sp$ is how many times faster the parallel run was, compared to the serial
baseline.
\end{definitionbox}

There are three regimes worth knowing:

\begin{itemize}[leftmargin=*]
  \item \textbf{Linear (ideal) speedup:} $\Sp = p$. Throwing $p$ resources at the problem reduced the time by exactly a factor of $p$.
  \item \textbf{Sub-linear speedup:} $\Sp < p$. The usual case. Sir listed the typical culprits: communication latency, synchronisation, cache coherence traffic, thread management, memory management, and the ``hundred small things'' that seep into the mixture.
  \item \textbf{Super-linear speedup:} $\Sp > p$. Rare but real. Usually it is a cache effect (the per-processor working set now fits in L2) or a search-parallelism effect (parallel exploration finds the answer earlier).
\end{itemize}

\subsection{Cost}

\begin{definitionbox}[Cost]
\[
\Cp \;=\; p \times \Tp
\]
The total computational effort: processor-seconds spent.
\end{definitionbox}

Speedup alone is dangerous. The slide gives a nice example: $\Tone = 100$\,s and
$T_8 = 20$\,s gives $S_8 = 5$, which sounds great. But the cost is $C_8 = 8 \times
20 = 160$ processor-seconds --- 60 more than the serial run did with one processor.
We got the answer faster, but we did $60\%$ more total work to get it. If energy
were our metric (\textbf{flops per joule}, the modern figure of merit for
\textit{green supercomputers}), this parallel run is actually worse.

\subsubsection*{Cost-optimality}
A parallel algorithm is \textbf{cost-optimal} when
\[
p \cdot \Tp \;=\; \mathcal{O}(\Tone),
\]
i.e.\ the parallel total work is asymptotically the same as the sequential work.
Sir gave a rather brutal counter-example to make the point:

\begin{examplebox}[Why cost matters: matrix multiplication]
Take $n \times n$ matrix multiplication, where the serial algorithm takes
$\Tone = \Theta(n^{3})$. Suppose we throw $p = n^{2}$ processors at it and somehow
get $\Tp = n$. Then
\[
\Sp = \frac{n^{3}}{n} = n^{2} \quad\text{(huge speedup),}
\qquad
\Cp = n^{2}\cdot n = n^{3}.
\]
This is fine --- cost-optimal. But suppose instead $\Tp = n^{2}$:
\[
\Sp = n, \qquad \Cp = n^{2}\cdot n^{2} = n^{4}.
\]
We pay $n^{4}$ total work for a speedup of only $n$. The cost has gone from $n^{3}$
to $n^{4}$, which is one full factor of $n$ of wasted work. As Sir put it, the
parallel solution is \textbf{unscalable} --- the speedup looks fine in isolation
but the bookkeeping reveals the trap.
\end{examplebox}

\subsection{Parallelism overhead}

\begin{definitionbox}[Total overhead]
\[
\To \;=\; \Cp - \Tone \;=\; p \cdot \Tp - \Tone
\]
\end{definitionbox}

This is the part of the parallel work that is not the serial algorithm's essential
computation. It bundles together:
\begin{itemize}[leftmargin=*]
  \item communication between processors (message passing, cache coherence traffic),
  \item synchronisation (barriers, locks, atomic updates, ordering fences),
  \item idle time from load imbalance,
  \item thread / process management (creation, scheduling, teardown),
  \item redundant computation done to avoid communication.
\end{itemize}

Cost-optimality is essentially the statement that the overhead does not grow
faster than $\Tone$: $\To/\Tone \to 0$ as the problem and $p$ scale together. If
$\To$ grows faster than $p$, scaling fails entirely --- adding processors only
makes things worse.

\subsection{Efficiency}

\begin{definitionbox}[Efficiency]
\[
\Ep \;=\; \frac{\Tone}{\Cp} \;=\; \frac{\Tone}{p \cdot \Tp} \;=\; \frac{\Sp}{p}
\]
The fraction of total processor-time spent doing useful work. $\Ep$ lies in
$(0,1]$, and $\Ep = 1$ is the ideal case.
\end{definitionbox}

The slide example is worth pausing on:

\begin{examplebox}[Speedup hides what efficiency reveals]
\begin{itemize}[leftmargin=*]
  \item $S_{8} = 7$ gives $E_{8} = 7/8 = 0.875$. Processors are 87.5\% utilised --- excellent.
  \item $S_{64} = 20$ gives $E_{64} = 20/64 \approx 0.3125$. On 64 processors, only 31\% of the aggregate capacity is doing useful work.
\end{itemize}
$S_{64}$ is almost three times $S_8$, but the 8-processor configuration is
twice as efficient. Speedup numbers reported on their own are half-truths; always
look at efficiency next to them.

The natural follow-up question --- \emph{if we want to keep efficiency at $0.875$
even on 64 processors, how much do we need to grow the problem?} --- is what
iso-efficiency answers in \S\ref{sec:isoefficiency}.
\end{examplebox}

% ==========================================================
\section{Scalability}
% ==========================================================

A speedup number is reported for one specific (problem size, processor count)
pair, which is why it is so often misleading. Scalability asks how the metrics
\emph{evolve} when we change either knob, or both.

From here on we treat $T$, $S$, $E$, $C$ as functions of two variables: the
problem size $n$ and the processor count $p$.

The reason this matters is simple. If we hold $n$ fixed and keep increasing $p$,
the per-processor work shrinks, but the communication and synchronisation costs
do not, so the speedup curve flattens out. If we instead grow $n$ together with
$p$, the parallel \emph{opportunity} grows --- there is more work to spread across
the processors --- while the serial portion of the program stays roughly the same
size in absolute terms. The two situations behave very differently, which is why
they get separate names.

\subsection{Strong scaling}

\begin{definitionbox}[Strong scaling]
Hold $n$ fixed and let $p$ grow. Study how $T(n,p)$ behaves.
Ideal strong scaling:
\[
T(n,p) \;=\; \frac{T(n,1)}{p}, \qquad \text{equivalently } S_p = p.
\]
\end{definitionbox}

Strong scaling rarely holds in practice for large $p$. Once the per-processor work
shrinks enough, the overheads dominate. A useful way to write the parallel runtime
is
\[
T(n,p) \;=\; T_{\text{comp}}(n,p) \;+\; T_{\text{comm}}(n,p) \;+\; T_{\text{sync}}(n,p) \;+\; T_{\text{idle}}(n,p).
\]
For fixed $n$, $T_{\text{comp}}$ falls like $1/p$, but the others typically stay
constant or grow with $p$, so eventually the second group dominates the first.
On top of this, load imbalance gets harder to avoid as the chunks get smaller, so
beyond some point adding more processors actively slows the program down.

\begin{errorbox}[Slide correction]
The slide writes ``Strong scaling breaks down: $T(n,p) = T_{\text{comp}} +
T_{\text{comm}} + T_{\text{sync}} + T_{\text{idle}}$.'' Sir corrected this in
class. The decomposition is the \emph{general} formula for parallel runtime --- it
is always true. What actually causes strong scaling to fail is that the
\emph{overhead terms} ($T_{\text{comm}} + T_{\text{sync}} + T_{\text{idle}}$) grow
with $p$. Without a lower bound on those, throwing more processors stops helping.
\end{errorbox}

\subsection{Weak scaling}

\begin{definitionbox}[Weak scaling]
Each processor is given a fixed workload $m$, so the total problem size is $n = mp$.
Ideal weak scaling:
\[
T(mp,\,p) \;=\; T(m,\,1),
\]
i.e.\ runtime stays constant as $p$ and $n$ grow together.
\end{definitionbox}

\begin{errorbox}[Slide correction]
The slide states ideal weak scaling as $T(mp, p) = T(m,1)/p$. This is wrong ---
that formula is what we would write for ideal \emph{strong} scaling on a problem
of size $m$. The whole point of weak scaling is that each processor's local work
stays the same, so the wall-clock time should stay constant, not shrink. The
correct formula is
\[
T(mp,p) = T(m,1) \quad\text{(or at worst approximately constant in $p$).}
\]
\end{errorbox}

\begin{examplebox}[Slide example]
\begin{center}
\begin{tabular}{@{}ccc@{}}
\toprule
$p$ & Total problem size & Time (s) \\ \midrule
1  & 1\,000\,000 & 10 \\
2  & 2\,000\,000 & 11 \\
4  & 4\,000\,000 & 13 \\ \bottomrule
\end{tabular}
\end{center}
Ideal weak scaling would have held the time at 10\,s. The drift from 10 to 11 to
13\,s is real communication overhead, but it is small and the growth is far
sub-linear in $p$. This is the kind of ``approximately constant'' behaviour Sir
called acceptable.
\end{examplebox}

% ==========================================================
\section{Amdahl's law}
% ==========================================================

The setup: split the sequential time $\Tone$ into
\begin{itemize}[leftmargin=*]
  \item a fraction $f$ that is inherently sequential and cannot be parallelised, and
  \item a fraction $(1-f)$ that can be parallelised perfectly.
\end{itemize}

On $p$ processors, the sequential part still costs $f \cdot \Tone$, while the
parallel part costs $(1-f)\cdot \Tone / p$. Adding them up:

\begin{definitionbox}[Amdahl's law]
\[
\boxed{\;\Tp \;=\; \Tone\!\left(f + \tfrac{1-f}{p}\right)\;}
\qquad
\boxed{\;\Sp \;=\; \dfrac{1}{f + \tfrac{1-f}{p}}\;}
\]
and as $p \to \infty$:
\[
\boxed{\;S_{\infty} \;=\; \frac{1}{f}.\;}
\]
\end{definitionbox}

\begin{errorbox}[Slide correction]
The slide writes
\[
T_p \;=\; T_1\bigl(1 + 1 - f/p\bigr),
\]
which is just garbled --- the parentheses and signs do not mean what they should.
The correct expression is
\[
T_p \;=\; T_1\!\left(f + \frac{1-f}{p}\right),
\]
which is consistent with the (correct) $S_p = 1/(f + (1-f)/p)$ printed
right below it.
\end{errorbox}

\subsection{What Amdahl's law actually tells us}

For a fixed-size problem, the serial fraction sets a hard ceiling on how fast you
can ever go, no matter how much hardware you throw at it.

\begin{itemize}[leftmargin=*]
  \item $f = 0.10$ (10\% serial) gives $S_\infty = 10$. This is the example on the slide.
  \item $f = 0.01$ (1\% serial) gives $S_\infty = 100$.
  \item $f = 0.001$ (0.1\% serial) gives $S_\infty = 1000$.
\end{itemize}

The lesson is that the serial fraction governs everything. Speeding up code that
is already parallel has diminishing returns; even a tiny reduction in the serial
portion can raise the ceiling dramatically.

\begin{insightbox}[The NVIDIA example Sir gave]
Imagine you are a product manager at NVIDIA with a \$10\,M budget for next quarter
and a known workload whose serial fraction you can estimate. The capacity-planning
calculation \emph{is} Amdahl's law: compute $S_\infty = 1/f$, identify the $p$ at
which the curve plateaus, and that is roughly the maximum number of GPUs worth
buying. Beyond that, you are paying for cards that cannot make the job any
faster.
\end{insightbox}

\subsection{What the speedup curves look like}

\begin{center}
\begin{tikzpicture}
\begin{axis}[
  width=0.85\linewidth, height=7cm,
  xlabel={Number of processors $p$},
  ylabel={Speedup $S_p$},
  xmin=1, xmax=64, ymin=0, ymax=22,
  grid=both, grid style={gray!20},
  legend pos=north west,
  legend style={font=\small, draw=primary},
  every axis plot/.append style={thick},
  title={Amdahl's law for several values of $f$}
]
\addplot [domain=1:64, samples=200, color=primary]
  {1/(0.05 + 0.95/x)};
\addlegendentry{$f=0.05$ \;($S_\infty=20$)}

\addplot [domain=1:64, samples=200, color=accent]
  {1/(0.10 + 0.90/x)};
\addlegendentry{$f=0.10$ \;($S_\infty=10$)}

\addplot [domain=1:64, samples=200, color=emerald]
  {1/(0.25 + 0.75/x)};
\addlegendentry{$f=0.25$ \;($S_\infty=4$)}

\addplot [domain=1:64, samples=200, color=amber]
  {1/(0.50 + 0.50/x)};
\addlegendentry{$f=0.50$ \;($S_\infty=2$)}
\end{axis}
\end{tikzpicture}
\captionof{figure}{Each curve plateaus at $1/f$ no matter how many processors are added.}
\end{center}

% ==========================================================
\section{Gustafson's law}
% ==========================================================

Amdahl's law assumes the problem size is fixed, which is rarely how things work in
practice. Most of the time we buy more processors precisely because we want to
solve \emph{bigger} problems --- not the same one slightly faster. Gustafson noticed
that in this scaled-workload setting, the serial portion of a program tends to
stay roughly the same in absolute terms, while the parallel portion grows with
$p$. So the effective serial fraction \emph{shrinks} as we scale up, and the
ceiling becomes much more forgiving.

\subsection{Setup}

Take the $p$-processor run as the reference, with runtime $T(n,p)$, and split it
as
\begin{itemize}[leftmargin=*]
  \item $f \cdot T(n,p)$ for the sequential portion,
  \item $(1-f)\cdot T(n,p)$ for the parallel portion.
\end{itemize}

If we ran the same workload on a single processor, the parallel portion would
take $p$ times longer (one processor doing what $p$ did) and the serial portion
would be unchanged. So
\[
T(n,1) \;=\; f\cdot T(n,p) \;+\; p \cdot (1-f) \cdot T(n,p).
\]

\begin{definitionbox}[Gustafson's law]
\[
\boxed{\;\Sp \;=\; \frac{T(n,1)}{T(n,p)} \;=\; f + (1-f)\cdot p \;=\; p - f(p-1)\;}
\]
\end{definitionbox}

This grows linearly in $p$, with no $1/f$ ceiling.

\begin{errorbox}[A subtle point: the $f$ in Gustafson is not the $f$ in Amdahl]
The slide reuses the symbol $f$ for both laws, which trips students up every year
and was the subject of a long Q\&A in this lecture. In Amdahl's law, $f$ is the
serial fraction of the \emph{serial} runtime $\Tone$. In Gustafson's law, $f$ is
the serial fraction of the \emph{parallel} runtime $T(n,p)$. They are different
numbers with different physical meanings, and treating them as the same is the
source of the question below.

Some textbooks rename Gustafson's $f$ to $s$ to avoid the collision; we will
mention this where it helps.
\end{errorbox}

\subsection{If Amdahl Limits Speedup, How Does Gustafson Avoid It?}
\label{sec:reconcile}

A natural question, asked by Ambar in class: if Amdahl says the speedup is capped
at $1/f$, how can Gustafson promise unbounded speedup for the same program?

\begin{qabox}[Reconciling the two laws]
If the two $f$'s were really the same number, equating the two speedup formulas
would force
\[
\frac{1}{f + (1-f)/p} \;\stackrel{?}{=}\; f + (1-f)p,
\]
which is plainly false --- the left side is bounded by $1/f$, the right side
grows without bound.

The way out is that Gustafson's $f$ is the serial fraction of the runtime on
$p$ processors, and that fraction \emph{depends on the problem size} even though
$f$ doesn't have $n$ written in it. As $n$ grows, the sequential portion stays
roughly the same in absolute time while the parallel portion grows, so the
proportion $f$ shrinks. Amdahl's $f$ does not shrink, because it is tied to the
fixed-size serial run.

So the two laws answer two different questions:
\begin{itemize}[leftmargin=*]
  \item \textit{Amdahl}: this exact problem, more processors --- how fast?
  \item \textit{Gustafson}: more processors and a proportionally bigger problem --- how does the bigger problem on $p$ processors compare to the same bigger problem on one processor?
\end{itemize}
\end{qabox}

\subsection{Reading Gustafson's law}

\begin{itemize}[leftmargin=*]
  \item If $f = 0.05$ on the parallel run and $p = 100$, Gustafson gives $\Sp = 0.05 + 0.95 \times 100 = 95.05$. Amdahl with the same nominal $f = 0.05$ would have capped this at $1/0.05 = 20$.
  \item The form $\Sp = p - f(p-1)$ is illuminating: the speedup is ``almost $p$'' minus a small correction. As $p$ grows, the correction is dwarfed by the $p$ and the speedup keeps climbing.
\end{itemize}

\subsection{Side-by-side comparison}

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{@{}p{0.22\linewidth} p{0.36\linewidth} p{0.36\linewidth}@{}}
\toprule
 & \textbf{Amdahl} & \textbf{Gustafson} \\ \midrule
Problem size & Fixed & Scales with $p$ \\
Meaning of $f$ & Serial fraction of $T_1$ & Serial fraction of $T(n,p)$ \\
Depends on problem size? & No & Yes (implicitly, through $T(n,p)$) \\
Speedup formula & $\Sp = \dfrac{1}{f + (1-f)/p}$ & $\Sp = f + (1-f)p$ \\
Limit as $p\to\infty$ & $1/f$ & Linear in $p$, no ceiling \\
Governs & Strong scaling & Weak scaling \\
\bottomrule
\end{tabular}
\end{center}

\begin{center}
\begin{tikzpicture}
\begin{axis}[
  width=0.85\linewidth, height=7cm,
  xlabel={Number of processors $p$},
  ylabel={Speedup $S_p$},
  xmin=1, xmax=64, ymin=0, ymax=65,
  grid=both, grid style={gray!20},
  legend pos=north west,
  legend style={font=\small, draw=primary},
  every axis plot/.append style={thick},
  title={Amdahl vs.\ Gustafson at 5\% serial fraction}
]
\addplot [domain=1:64, samples=200, color=accent]
  {1/(0.05 + 0.95/x)};
\addlegendentry{Amdahl ($f=0.05$): ceiling $=20$}

\addplot [domain=1:64, samples=200, color=emerald]
  {0.05 + 0.95*x};
\addlegendentry{Gustafson ($f=0.05$): linear}

\addplot [domain=1:64, samples=2, color=gray, dashed]
  {x};
\addlegendentry{Ideal $S_p = p$}
\end{axis}
\end{tikzpicture}
\captionof{figure}{Same nominal $5\%$ serial fraction, two very different
predictions. They disagree because they are not really talking about the same
quantity (see the discussion in \S\ref{sec:reconcile}).}
\end{center}

% ==========================================================
\section{Iso-efficiency: how fast must the problem grow?}
\label{sec:isoefficiency}
% ==========================================================

Sir closed the lecture by going back to the question that came up when we first
defined efficiency. We saw that $E_{64} = 0.3125$ was much worse than $E_8 = 0.875$
for the same program. The natural inverse question is:

\begin{insightbox}[The iso-efficiency question]
At what rate must the problem size $n$ grow with the processor count $p$ so that
$E(n, p)$ stays at some chosen constant $k$?
\end{insightbox}

This is the iso-efficiency view. It pulls together everything we have done:
efficiency is the quality, $n$ and $p$ are the controls, and we ask how the
controls must move together.

\subsection{Derivation}

Start from the definition,
\[
E(n, p) \;=\; \frac{T(n,1)}{p \cdot T(n,p)}.
\]
Use $\To = p \cdot T(n,p) - T(n,1)$ to rewrite $T(n,p) = (T(n,1) + \To)/p$, which
gives
\[
E(n,p) \;=\; \frac{1}{1 + \To(n,p)/T(n,1)}.
\]
If we want this to equal a target efficiency $k < 1$, we solve for $T(n,1)$:
\[
T(n,1) \;=\; \frac{k}{1-k}\cdot \To(n,p).
\]
$T(n,1)$ is essentially the work $W(n)$ done by the serial algorithm, so we can
write
\[
\boxed{\;W(p) \;=\; \frac{k}{1-k}\cdot \To(n,p)\;}
\]
This is the \textbf{iso-efficiency function}: it tells us how much work the
problem must contain to absorb the overhead and keep $E$ at $k$.

\subsection{Strongly scalable vs.\ weakly scalable}

The growth rate of $W(p)$ summarises everything about an algorithm's scalability.
The classification Sir gave verbally:

\begin{definitionbox}[Strongly vs.\ weakly scalable]
Let $W(p)$ be the problem size needed to hold efficiency at a chosen $k$.
\begin{itemize}[leftmargin=*]
  \item If $W(p)$ grows \emph{slowly} with $p$ --- constant, logarithmic, or at worst a modest polynomial --- the algorithm is \textbf{strongly scalable}.
  \item If $W(p)$ has to grow as fast as $p$ (or faster) just to hold efficiency steady, the algorithm is only \textbf{weakly scalable}.
\end{itemize}
\end{definitionbox}

\begin{errorbox}[Don't confuse the two ``strong / weak'' meanings]
``Strong scaling'' and ``weak scaling'' (Section 4) describe \emph{experimental
regimes} --- whether you hold $n$ fixed or grow it with $p$ when you run your
benchmark. ``Strongly scalable'' and ``weakly scalable'' here describe an
\emph{algorithmic property} --- how fast its iso-efficiency function grows. Same
words, different things; pay attention to context. Different textbooks also use
``scalable parallel algorithm'' in slightly different ways. The iso-efficiency
function itself is unambiguous, so when in doubt, just write it down.
\end{errorbox}

\subsection{Some rules of thumb}

\begin{itemize}[leftmargin=*]
  \item $W(p) = \Theta(p)$ --- good. Doubling processors needs doubling the work to keep efficiency constant. Many dense linear algebra routines sit here.
  \item $W(p) = \Theta(p \log p)$ --- still respectable. Typical of parallel sorts and prefix-sum algorithms.
  \item $W(p) = \Theta(p^2)$ or worse --- poor. The problem has to blow up quadratically just to stand still on efficiency. Algorithms in this category are usually redesigned or avoided at large $p$.
\end{itemize}

This also closes the loop on the example from earlier: if we want the 64-processor
run to keep the $E_8 = 0.875$ efficiency we measured at $p=8$, the iso-efficiency
function tells us exactly how much we need to grow $n$. A library writer who
hands users a fixed implementation without telling them this growth rate is, in
Sir's word, selling a ``facade'' of parallelism.

% ==========================================================
\section{A few practical points}
% ==========================================================

The two classical laws are clean but they paper over a few things that matter in
real systems. None of these were on the slides; they are standard things worth
keeping in mind.

\subsection{Communication-aware models}

Both Amdahl and Gustafson assume the parallel portion scales perfectly. Real code
does not. A simple refinement is
\[
T(n,p) \;=\; \frac{W(n)}{p} \;+\; T_{\text{comm}}(n,p),
\]
where $T_{\text{comm}}$ is typically of the form $\alpha + \beta \cdot m$ for a
message of $m$ bytes, $\alpha$ being the latency and $\beta$ the inverse
bandwidth. In many distributed settings $T_{\text{comm}}$ grows like $\log p$ or
$\sqrt{p}$, which slowly eats into the speedup as $p$ grows.

\subsection{Super-linear speedup is real}

The textbook definitions rule out $\Sp > p$, but it does happen, usually for one
of three reasons:
\begin{itemize}[leftmargin=*]
  \item \textbf{Cache:} the partitioned working set fits in faster memory than the original did.
  \item \textbf{Search:} parallel branch-and-bound prunes the tree earlier.
  \item \textbf{Memory bandwidth:} more nodes mean more aggregate memory bandwidth, not just more cores.
\end{itemize}

\subsection{Energy and the green supercomputer}

Sir mentioned the rise of the ``green supercomputer'' movement from his early
career. Once GPGPUs entered HPC, raw FLOPS stopped being a sufficient figure of
merit --- two systems with the same FLOPS could burn very different amounts of
electricity. The metric people switched to is \textbf{flops per joule}: useful
computation per unit of energy. Looked at this way, the cost example we did
earlier ($\Cp = 160$ vs.\ $\Tone = 100$) is not just a wall-clock story but a
$60\%$ extra energy bill for the same answer.

\subsection{What to actually report}

\begin{insightbox}[What to report when benchmarking a parallel program]
\begin{enumerate}[leftmargin=*]
  \item $\Tone$ (best serial) and $\Tp$ for several $p$.
  \item Both strong-scaling and weak-scaling curves where they make sense.
  \item Efficiency $\Ep$, not just speedup --- it is where the wasted work shows up.
  \item A breakdown of $\Tp$ into computation, communication, synchronisation, idle.
  \item The iso-efficiency function (or an empirical estimate) if the algorithm is intended for large $p$.
  \item Problem size: without $n$, the numbers don't mean anything.
\end{enumerate}
\end{insightbox}

% ==========================================================
\section{Summary}
% ==========================================================

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{@{}l l l@{}}
\toprule
\textbf{Quantity} & \textbf{Definition} & \textbf{Ideal value} \\ \midrule
Speedup $\Sp$ & $\Tone / \Tp$ & $p$ \\
Cost $\Cp$ & $p \cdot \Tp$ & $\Theta(\Tone)$ \\
Overhead $\To$ & $\Cp - \Tone$ & $\to 0$ asymptotically \\
Efficiency $\Ep$ & $\Tone / (p \cdot \Tp) = \Sp / p$ & $1$ \\
Strong scaling & $T(n,p)$ for fixed $n$ & $T(n,1)/p$ \\
Weak scaling & $T(mp,p)$ for fixed $m$ & $T(m,1)$ \\
Amdahl & $\Sp = 1/(f + (1-f)/p)$ & $\to 1/f$ \\
Gustafson & $\Sp = f + (1-f)p$ & linear in $p$ \\
Iso-efficiency $W(p)$ & work needed to hold $E = k$ & slow-growing in $p$ \\
\bottomrule
\end{tabular}
\end{center}

\begin{insightbox}[Things to take away]
\begin{enumerate}[leftmargin=*]
  \item Speedup without efficiency lies. $S_{64} = 20$ sounds great until you notice $E_{64} = 0.31$.
  \item Strong and weak scaling are different experiments and tell different stories. Report both when both make sense.
  \item Amdahl and Gustafson don't actually contradict each other --- they are answering different questions, and their $f$'s mean different things.
  \item The serial fraction is the enemy. Tiny serial sections dominate at large $p$.
  \item Overheads are not a footnote. Communication, synchronisation, and idle time are where most speedups die.
  \item Iso-efficiency is the honest scalability metric --- it forces you to say ``how fast at what size?''
\end{enumerate}
\end{insightbox}

\vfill
\begin{center}
\rule{0.6\linewidth}{0.4pt}\\[0.3em]
{\small\itshape End of Lecture 21 notes}
\end{center}

\end{document}
